Inspiring Confidence and Easing Insecurity in the Pre-election period: The Effect of Recruiting Military Veterans on Confidence in 2024 US Elections

### 2025-03-19 docs/_tex/index.tex
clean repo and push to publish

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
    \setmainfont[]{TeX Gyre Pagella}
  \setmathfont[]{TeX Gyre Pagella Math}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

@@

\maketitle
%\noindent \rule{\linewidth}{.5pt}
%\noindent \rule{\linewidth}{.5pt}

@@

In this paper, I report results from a recent survey experiment
administered to test whether publicized efforts to recruit veterans to
work as election staff and volunteers would improve public trust in
elections and ease election insecurity. Results of the survey experiment
support the notion that emphasizing veterans as the target of election
worker recruitment efforts eases pre-election insecurity. Expectations
of electoral fraud and concerns for voter safety were lower among those
who read an announcement that veterans are being recruited to work as
election staff and volunteers compared to those who read a control
vignette where veterans were not mentioned\footnote{The vignettes
  written were based on real news stories}. Notable is that there was a
significant difference in confidence among those in the treatment
condition who believe that results of the 2020 election were
illegitimate.

@@

Intuitively, enhancing public trust in elections would best be
accomplished by enhancing the \emph{trustworthiness} of the institution,
i.e., consistently doing the things that election officials already
regularly do come election time. However, trust and confidence in
elections has become ever more precarious over the last few election
cycles. Especially considering public polling data since 2000 shows that
confidence that votes were, or would be, counted as intended was in a
consistent decline despite efforts towards bolstering election integrity
and trustworthiness (\citeproc{ref-sances2015}{Sances and Stewart
2015}). This is even more pronounced considering the role that
partisanship has had on such confidence over accuracy of vote count
(\citeproc{ref-sances2015}{Sances and Stewart 2015};
\citeproc{ref-stewart2022}{Stewart 2022}).

@@

As such, it is questionable whether we can characterize public trust in
the pre-election period as the same trust after the election results
have come out. The former is \emph{anticipatory}---i.e., the kind that
is more or less anxious given the uncertainties surrounding the
election. The latter is \emph{empirical}---a judgement discerned in
hindsight after the experience of the election event has occurred. As
discussed by Stewart (\citeproc{ref-stewart2022}{2022}), not only was
confidence in vote count accuracy influenced by the voter experience,
but so too were evaluations of election officials
(\citeproc{ref-stewart2022}{2022, 242--43}). Not to mention the
influence that the election results would also have on such evaluations.
In other words, trust evaluations post-election are influenced by one's
interpretation of their experience and election results. This study, in
contrast, focuses primarily on that \emph{anticipatory} kind of
confidence, which speaks more to those insecurities based on perceptions
of the institution's trustworthiness than upon the particular voter
experience.

@@

Election officials are likely to be agnostic as to who dedicates their
time to civil service such as election work. Staffing issues have been
an issue since at least the 1990s (\citeproc{ref-ferrer2024}{Ferrer,
Thompson, and Orey 2024}; \citeproc{ref-maidenberg1996}{Maidenberg
1996}). In 2020, such efforts were made far more difficult by the
COVID-19 pandemic (\citeproc{ref-abbate2020a}{Abbate 2020};
\citeproc{ref-mena2020}{Mena 2020}). There's no special reason to target
veterans for recruitment above other groups. Indeed, there's no reason
to discriminate recruitment efforts at all if the point is purely to
fill staffing vacancies.

@@

It should be noted that Maricopa County, AZ was chosen as the setting of
the story in the vignette due to the increased scrutiny levied toward
election administration there after the 2020 election
(\citeproc{ref-giles2021}{Giles 2021};
\citeproc{ref-maricopacountyelectionsdepartment2022}{Maricopa County
Elections Department 2022}). Because of this, treatment effects are
potentially limited or constrained to attitudes concerning the location
specified in the vignette. A block of survey items asked specifically
about Maricopa County, AZ, followed by an identical block of items that
asked the same questions about one's local area. Later, I compare and
discuss results between the items pertaining to different settings.

@@

The survey included items that assessed both trust in elections
administration and expectations of electoral fraud, referred to simply
as \emph{trust} and \emph{distrust}, respectively. Responses from these
two sets of items were utilized to construct the primary dependent
variable, \emph{confidence in elections}.

Prior to the reading the treatment or control vignette, participants
responded to distinct survey items assessing one's favorability to local
election officials (``In general, how favorable or unfavorable is your
impression of local election officials?'') and legitimacy of the 2020
election results (``Regardless of whom you supported in the 2020
election, do you think Joe Biden's election as president was legitimate,
or was he not legitimately elected?''). The former presented five
response options from ``Strongly favorable'' to ``Strongly
unfavorable''; the latter item presented a binary of either
``Legitimate'' or ``Not Legitimate''.

@@

Finally, a pair of survey items assessed concerns for potential violence
and confidence in voter safety. The pair was asked once in regard to
Maricopa County, AZ and again with respect to one's local
area\footnote{Survey item wording and responses are included in Appendix
  A.}.

>>>>

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
    \setmainfont[]{TeX Gyre Pagella}
    \setsansfont[]{TeX Gyre Pagella}
  \setmathfont[]{TeX Gyre Pagella Math}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

@@

\maketitle
%\noindent \rule{\linewidth}{.5pt}
\begin{abstract}
In this paper, I report results from a recent survey experiment
administered to test whether publicized efforts to recruit veterans to
work as election staff and volunteers would improve public trust in
elections and ease election insecurity. Results of the survey experiment
support the notion that emphasizing veterans as the target of election
worker recruitment efforts eases pre-election insecurity. Expectations
of electoral fraud and concerns for voter safety were lower among those
who read an announcement that veterans are being recruited to work as
election staff and volunteers compared to those who read a control
vignette where veterans were not mentioned. Notable is that there was a
significant difference in confidence among those in the treatment
condition who believe that results of the 2020 election were
illegitimate. In addition, concerns over voter safety and potential for
violence were lower among those who received the treatment.
\end{abstract}
%\noindent \rule{\linewidth}{.5pt}

@@

In this paper, I report results from a recent survey experiment
administered to test whether publicized efforts to recruit veterans to
work as election staff and volunteers would improve public trust in
elections and ease election insecurity. Results of the survey experiment
support the notion that emphasizing veterans as the target of election
worker recruitment efforts eases pre-election insecurity. Expectations
of electoral fraud and concerns for voter safety were lower among those
who read an announcement that veterans are being recruited to work as
election staff and volunteers compared to those who read a control
vignette where veterans were not mentioned\footnote{The vignettes
  written were based on real news stories}. Notable is that there was a
significant difference in confidence among those in the treatment
condition who believe that results of the 2020 election were
illegitimate. In addition, concerns over voter safety and potential for
violence were lower among those who received the treatment.

@@

Intuitively, enhancing public trust in elections would best be
accomplished by enhancing the \emph{trustworthiness} of the institution,
e.g., enhancing vote tabulation equipment and technology, conducting
audits. However, trust and confidence in elections has become ever more
precarious over the last few election cycles. Especially considering
public polling data since 2000 shows that confidence that votes were, or
would be, counted as intended was in a consistent decline despite
efforts towards bolstering election integrity and trustworthiness
(\citeproc{ref-sances2015}{Sances and Stewart 2015}). This is even more
pronounced considering the role that partisanship has had on such
confidence over accuracy of vote counts
(\citeproc{ref-sances2015}{Sances and Stewart 2015};
\citeproc{ref-stewart2022}{Stewart 2022}).

@@

As such, it is questionable whether we can characterize public trust in
the pre-election period as the same trust after the election results
have come out. The former is \emph{anticipatory}---i.e., the kind that
is more or less anxious given the uncertainties surrounding the
election. The latter is \emph{empirical}---a judgement discerned in
hindsight after the experience of the election event has occurred. As
discussed by Stewart (\citeproc{ref-stewart2022}{2022}), not only was
confidence in vote count accuracy influenced by the voter experience,
but so too were evaluations of election officials
(\citeproc{ref-stewart2022}{2022, 242--43}). Not to mention the
influence that the election results would also have on such evaluations.
In other words, evaluations of trust considered in hindsight are
influenced by the voter experience and the outcome of the election
results. This study, in contrast, focuses primarily on that
\emph{anticipatory} kind of confidence, which speaks more to those
insecurities based on perceptions of the institution's trustworthiness
than upon the particular voter experience.

@@

Election officials are likely to be agnostic as to who dedicates their
time to civil service such as election work. Especially since staffing
issues have been an issue since at least the 1990s
(\citeproc{ref-ferrer2024}{Ferrer, Thompson, and Orey 2024};
\citeproc{ref-maidenberg1996}{Maidenberg 1996}). In 2020, such efforts
were made far more difficult by the COVID-19 pandemic
(\citeproc{ref-abbate2020a}{Abbate 2020}; \citeproc{ref-mena2020}{Mena
2020}). There's no special reason to target veterans for recruitment
above other groups. Indeed, there's no reason to discriminate
recruitment efforts at all if the point is purely to fill staffing
vacancies.

@@

It should be noted that Maricopa County, AZ was chosen as the setting of
the story in the vignette due to the increased scrutiny levied toward
election administration there after the 2020 election
(\citeproc{ref-giles2021}{Giles 2021};
\citeproc{ref-maricopacountyelectionsdepartment2022}{Maricopa County
Elections Department 2022}). Because of this, treatment effects are
potentially limited or constrained to attitudes concerning the location
specified in the vignette. A block of survey items asked specifically
about Maricopa County, AZ, followed by an identical block of items that
asked the same questions about one's local area. I compare and discuss
results between the items pertaining to different settings.

@@

The survey included items that assessed both trust in elections
administration and expectations of electoral fraud, referred to simply
as \emph{trust} and \emph{distrust}, respectively. Responses from these
two sets of items were utilized to construct the primary dependent
variable, \emph{confidence in elections}. Additionally, a pair of survey
items assessed concerns for potential violence and confidence in voter
safety. The pair was asked once in regard to Maricopa County, AZ and
again with respect to one's local area\footnote{Question wording and
  responses to a selection of survey items are included in Appendix A.}.
Although two survey items are insufficient to serve as a composite
scale, safety concerns will be evaluated by examination of response to
these two questions in particular.

Prior to the reading the treatment or control vignette, participants
responded to distinct survey items assessing one's beliefs about the
legitimacy of the 2020 election results (``Regardless of whom you
supported in the 2020 election, do you think Joe Biden's election as
president was legitimate, or was he not legitimately elected?'').
Respondents selected either ``Legitimate'' or ``Not Legitimate''.

@@
